noam_data_exploration_v1

### 2024-11-20 Personal_Prelim_Code/mcr_algal_data_cleaning_final.Rmd

# Caulerpa peltata is outdated --> Caulerpa racemosa
data$Taxonomy_Substrate_Functional_Group[data$Taxonomy_Substrate_Functional_Group == "Caulerpa peltata"] <- "Caulerpa racemosa"

>>>>

# Caulerpa peltata is outdated --> Caulerpa racemosa
data$Taxonomy_Substrate_Functional_Group[data$Taxonomy_Substrate_Functional_Group == "Caulerpa peltata"] <- "Caulerpa chemnitzia"

### 2024-11-20 Personal_Prelim_Code/noam_data_exploration_v3.Rmd

```{r import and format data}
# starting with all data now because of 0 algae transects/quadrats
data <- read.csv(here::here("data", "allbenthicdata_cleaned_colnames.csv"), stringsAsFactors = F)

not_algae <- c("ascidian", "bare_space", "coral", "coral_rubble", "corallimorpharia", "heteractis_sp", "millepora_platyphylla", "no_data", "sand", "sacrophyton_sp", "shell_debris", "soft_coral", "sponge", "tridacna_sp", "zooxanthellae", "row_sums")

@@

meta_cols <- c(colnames(data[1:7] ), "island_side", "island_side_habitat", "stratum")
species_cols <- colnames(data)[!colnames(data) %in% meta_cols]

#pivot long 
long_data <- pivot_longer(
  data = data,
  cols = 8:77,
  names_to = "taxa",
  values_to = "percent_cover")

@@

  # fit gls for current time series
  loop_mod <- gls(logRich ~ year, correlation =corAR1(form=~year), 
                  control=glsControl(msVerbose=T), data = newData)

@@

summary(richness_mod_2) # looks like strong autocorrelation, which makes sense to me
Anova(richness_mod_2)
plot(allEffects(richness_mod_2))
mod_emm <- emmeans(richness_mod_2, ~ habitat * island_side * year_factor) # if we want to compare habitats year to year
richness_plot_info <- plot(mod_emm, plotit = FALSE)
richness_plot_info$year_numeric <- as.numeric(levels(richness_plot_info$year_factor))
(emmeans_graph <- 
  ggplot(data =richness_plot_info) +
  geom_point(aes(y = exp(the.emmean), x = year_numeric, colour = island_side), size = 8) +
  geom_errorbar(aes(ymin = exp(asymp.LCL), ymax = exp(asymp.UCL), x = year_numeric, colour = island_side), 
                width = 0.2, size = 1.5) +
  geom_segment(aes(y = 5.2, yend = 5.2, x = 2006, xend = 2010), linewidth = 4, alpha = 0.2, colour = "grey") +
  geom_vline(xintercept = 2010, linetype = 2, colour = "black", size = 2) +
  geom_vline(xintercept = 2019, linetype = 2, colour = "red", size = 2) +
  scale_y_log10() +
  facet_wrap(~habitat) +
  theme_classic() +
  labs(x = "Year", y = "Estimated marginal mean richness\n") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
#        panel.background = element_blank(),
        legend.text = element_text(size=20),
        legend.title = element_text(size=20),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        axis.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        strip.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain") ) )

@@

# Julianna's model structure
j_rich_mod <- glmmTMB(richness ~ year + habitat + (1|site/transect),
               family = poisson,
               data = alpha_diversity_quad)
summary(j_rich_mod)
plot(allEffects(j_rich_mod))

grid <- expand_grid(year = unique(alpha_diversity_quad$year),
                       habitat = unique(alpha_diversity_quad$habitat),
                       site = unique(alpha_diversity_quad$site),
                       island_side = unique(alpha_diversity_quad$island_side),
                       transect = unique(alpha_diversity_quad$transect))
grid$Preds <- predict(j_rich_mod, newdata = grid)

grid %>% 
  # get transect means
  group_by(year, habitat, island_side) %>% 
  summarize(Pred_mean_richness = mean(Preds),
            SD_richness = sd(Preds)) %>% 
  ggplot(aes(x = year, y = Pred_mean_richness, group = habitat, color = habitat)) +
  geom_line() +
  theme_bw() +
  facet_wrap(~island_side)

@@

summary(shannon_mod) # looks like strong autocorrelation, which makes sense to me
Anova(shannon_mod)
plot(allEffects(shannon_mod_2))
mod_emm <- emmeans(shannon_mod, ~ habitat * island_side * year_factor) # if we want to compare habitats year to year
shannon_plot_info <- plot(mod_emm, plotit = FALSE)
shannon_plot_info$year_numeric <- as.numeric(levels(shannon_plot_info$year_factor))
(emmeans_graph <- 
  ggplot(data =shannon_plot_info) +
  geom_point(aes(y = exp(the.emmean), x = year_numeric, colour = island_side), size = 8) +
  geom_errorbar(aes(ymin = exp(lower.CL), ymax = exp(upper.CL), x = year_numeric, colour = island_side), 
                width = 0.2, size = 1.5) +
  geom_segment(aes(y = 2.5, yend = 2.5, x = 2006, xend = 2010), linewidth = 4, alpha = 0.2, colour = "grey") +
  geom_vline(xintercept = 2010, linetype = 2, colour = "black", size = 2) +
  geom_vline(xintercept = 2019, linetype = 2, colour = "red", size = 2) +
  scale_y_log10() +
  facet_wrap(~habitat) +
  theme_classic() +
  labs(x = "Year", y = "Shannon-Weiner Index (H')\n") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
#        panel.background = element_blank(),
        legend.text = element_text(size=20),
        legend.title = element_text(size=20),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        axis.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        strip.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain") ) )

@@

```{r mean rank shift, fig.height = 24, fig.width=16}
# rank shift

@@

```{r rate of change, fig.height=15, fig.width=15}
# rate change
long_data_sub$percent_cover_transformed <- wisconsin(long_data_sub$percent_cover) # wisconsin transform abundance data

@@

```{r rate change vis method 2, fig.height=15, fig.width=15}
rate_change_untransformed<-
  rate_change_interval(
  df = long_data_sub,
  time.var = "year",
  species.var = "taxa",
  abundance.var = "percent_cover",
  replicate.var = "location"
)

@@

```{r CV, fig.height = 8, fig.width = 10}
# coefficient of variation
CV<-function(x){
  return(sd(x,na.rm=T)/mean(x,na.rm=T))
}

>>>>

```{r import and format data}
# coefficient of variation
CV<-function(x){
  return(sd(x,na.rm=T)/mean(x,na.rm=T))
}

# starting with all data now because of 0 algae transects/quadrats
data <- read.csv(here::here("data", "allbenthicdata_cleaned_colnames.csv"), stringsAsFactors = F)

not_algae <- c("ascidian", "bare_space", "coral", "coral_rubble", "corallimorpharia", "heteractis_sp", "millepora_platyphylla", "no_data", "sand", "sarcophyton_sp", "shell_debris", "soft_coral", "sponge", "tridacna_sp", "zooxanthellae", "row_sums")

@@

meta_cols <- c("year", "location", "site", "habitat", "site_habitat", "transect", "quadrat", "island_side", "island_side_habitat", "stratum")
species_cols <- colnames(data)[!colnames(data) %in% meta_cols]

#pivot long 
long_data <- pivot_longer(
  data = data,
  cols = 8:76,
  names_to = "taxa",
  values_to = "percent_cover")

@@

  # fit gls for current time series
  loop_mod <- gls(richness ~ year, correlation =corAR1(form=~year), 
                  control=glsControl(msVerbose=T), data = newData)

@@

summary(richness_mod_2) # looks like strong autocorrelation, which makes sense to me
Anova(richness_mod_2)
plot(allEffects(richness_mod_2))
mod_emm <- emmeans(richness_mod_2, ~ habitat * island_side * year_factor) # if we want to compare habitats year to year
richness_plot_info <- plot(mod_emm, plotit = FALSE)
richness_plot_info$year_numeric <- as.numeric(levels(richness_plot_info$year_factor))
(emmeans_graph <- 
  ggplot(data =richness_plot_info) +
  geom_point(aes(y = exp(the.emmean), x = year_numeric, colour = island_side), size = 8) +
  geom_errorbar(aes(ymin = exp(asymp.LCL), ymax = exp(asymp.UCL), x = year_numeric, colour = island_side), 
                width = 0.2, size = 1.5) +
  geom_segment(aes(y = 5.2, yend = 5.2, x = 2006, xend = 2010), linewidth = 4, alpha = 0.2, colour = "grey") +
  geom_vline(xintercept = 2010, linetype = 2, colour = "black", size = 2) +
  geom_vline(xintercept = 2019, linetype = 2, colour = "red", size = 2) +
  scale_y_log10() +
  facet_wrap(~habitat) +
  theme_classic() +
  labs(x = "Year", y = "Richness\n") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
#        panel.background = element_blank(),
        legend.text = element_text(size=20),
        legend.title = element_text(size=20),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        axis.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        strip.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain") ) )

@@

# Julianna's model structure
# j_rich_mod <- glmmTMB(richness ~ year + habitat + (1|site/transect),
#                family = poisson,
#                data = alpha_diversity_quad)
# summary(j_rich_mod)
# plot(allEffects(j_rich_mod))
# 
# grid <- expand_grid(year = unique(alpha_diversity_quad$year),
#                        habitat = unique(alpha_diversity_quad$habitat),
#                        site = unique(alpha_diversity_quad$site),
#                        island_side = unique(alpha_diversity_quad$island_side),
#                        transect = unique(alpha_diversity_quad$transect))
# grid$Preds <- predict(j_rich_mod, newdata = grid)
# 
# grid %>% 
#   # get transect means
#   group_by(year, habitat, island_side) %>% 
#   summarize(Pred_mean_richness = mean(Preds),
#             SD_richness = sd(Preds)) %>% 
#   ggplot(aes(x = year, y = Pred_mean_richness, group = habitat, color = habitat)) +
#   geom_line() +
#   theme_bw() +
#   facet_wrap(~island_side)

@@

summary(shannon_mod) # looks like strong autocorrelation, which makes sense to me
Anova(shannon_mod)

mod_emm <- emmeans(shannon_mod, ~ habitat * island_side * year_factor) # if we want to compare habitats year to year
shannon_plot_info <- plot(mod_emm, plotit = FALSE)
shannon_plot_info$year_numeric <- as.numeric(levels(shannon_plot_info$year_factor))
(emmeans_graph <- 
  ggplot(data =shannon_plot_info) +
  geom_point(aes(y = exp(the.emmean), x = year_numeric, colour = island_side), size = 8) +
  geom_errorbar(aes(ymin = exp(lower.CL), ymax = exp(upper.CL), x = year_numeric, colour = island_side), 
                width = 0.2, size = 1.5) +
  geom_segment(aes(y = 2.5, yend = 2.5, x = 2006, xend = 2010), linewidth = 4, alpha = 0.2, colour = "grey") +
  geom_vline(xintercept = 2010, linetype = 2, colour = "black", size = 2) +
  geom_vline(xintercept = 2019, linetype = 2, colour = "red", size = 2) +
  scale_y_log10() +
  facet_wrap(~habitat) +
  theme_classic() +
  labs(x = "Year", y = "Shannon-Weiner Index (H')\n") +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
#        panel.background = element_blank(),
        legend.text = element_text(size=20),
        legend.title = element_text(size=20),
        axis.line = element_line(colour = "black"),
        axis.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        axis.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = .5, face = "plain"),
        axis.title.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = 0, face = "plain"),
        strip.text.x = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain"),
        strip.text.y = element_text(color = "grey20", size = 25, hjust = .5, vjust = 0, face = "plain") ) )

@@

```{r mean rank shift, fig.height = 24, fig.width=16, eval = FALSE}
# rank shift

@@

```{r rate of change, fig.height=15, fig.width=15, eval = FALSE}
# rate change
long_data_sub$percent_cover_transformed <- wisconsin(long_data_sub$percent_cover) # wisconsin transform abundance data

@@

```{r rate change vis method 2, fig.height=15, fig.width=15, eval = FALSE}
rate_change_untransformed<-
  rate_change_interval(
  df = long_data_sub,
  time.var = "year",
  species.var = "taxa",
  abundance.var = "percent_cover",
  replicate.var = "location"
)

@@

```{r CV, fig.height = 8, fig.width = 10}
